---
title: "Kids ARC: IRT analysis"
subtitle: "Rasch (1PL) comparison of human and LLM performance"
author: "OJ"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: false
  message: true
---

```{r}
#| label: setup
#| include: false

# knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| label: libraries

library(mirt)
library(tidyr)
library(MASS)     # load before dplyr so dplyr::select isn't masked
library(dplyr)
library(ggplot2)
library(ggrepel)
library(ggExtra)
```

## Data

```{r}
#| label: load-data
df_human <- read.csv("../data/preprocessed/human_generation.csv")
df_llm   <- read.csv("../data/preprocessed/llm_generation.csv")
df_items <- read.csv("../data/preprocessed/items.csv")
df_items$size <- as.factor(df_items$size)
```

## Preprocessing functions

`simple_irt_preprocess()` averages `concept_response` within participant (or
model) and `item_main_id`, pivots to a wide item matrix, and reports items at
0% or 100% correct.

```{r}
#| label: fn-preprocess

simple_irt_preprocess <- function(df, type = c("human", "llm")) {
  type <- match.arg(type)
  id_col <- if (type == "human") "participant" else "model"

  df_avg <- df %>%
    dplyr::summarise(
      concept_response = mean(concept_response, na.rm = TRUE),
      .by = c(!!rlang::sym(id_col), item_main_id)
    )

  wide <- df_avg %>%
    dplyr::select(!!rlang::sym(id_col), item_main_id, concept_response) %>%
    tidyr::pivot_wider(names_from = item_main_id, values_from = concept_response)

  if (type == "human") {
    wide <- wide %>% dplyr::filter(participant != 0)
  }

  wide <- wide %>% dplyr::select(-!!rlang::sym(id_col))

  item_names <- colnames(wide)
  item_means <- colMeans(wide, na.rm = TRUE)

  all_0_items <- item_names[item_means == 0]
  all_1_items <- item_names[item_means == 1]
  bad_items <- c(all_0_items, all_1_items)

  if (length(all_0_items) > 0) {
    message("Items with 0% correct (possible scoring bug): ",
            paste(all_0_items, collapse = ", "))
  }
  if (length(all_1_items) > 0) {
    message("Items with 100% correct (possible scoring bug): ",
            paste(all_1_items, collapse = ", "))
  }

  list(wide = wide, bad_items = bad_items)
}
```

`find_constant_items()` catches items that collapse to a single response
category *after* rounding — these cannot be estimated by `mirt`.

```{r}
#| label: fn-constant-items

find_constant_items <- function(wide) {
  colnames(wide)[sapply(wide, function(x) length(unique(na.omit(x))) < 2)]
}
```

Shared cleaning step, so the human and LLM branches stay identical:

```{r}
#| label: fn-clean

clean_for_irt <- function(res, label) {
  wide <- res$wide %>% dplyr::select(-dplyr::any_of(res$bad_items))
  wide <- wide %>% dplyr::mutate(dplyr::across(dplyr::everything(), round))

  constant_items <- find_constant_items(wide)
  if (length(constant_items) > 0) {
    message(label, " items constant after rounding (dropped): ",
            paste(constant_items, collapse = ", "))
  }
  wide %>% dplyr::select(-dplyr::any_of(constant_items))
}
```

## Build item matrices

```{r}
#| label: preprocess-human

res_human  <- simple_irt_preprocess(df_human, type = "human")
wide_human <- clean_for_irt(res_human, "Human")

dim(wide_human)
```

```{r}
#| label: preprocess-llm

res_llm  <- simple_irt_preprocess(df_llm, type = "llm")
wide_llm <- clean_for_irt(res_llm, "LLM")

dim(wide_llm)
```

## Item accuracy (% correct)

::: callout-note
These means come from the **cleaned** matrices, so they describe the same item
set the Rasch models below are fit on. The original script computed them from
`res_*$wide` (before dropping bad and constant items), which meant the accuracy
plots and the IRT models were describing different item sets.
:::

```{r}
#| label: item-means

item_means_human <- colMeans(res_human$wide, na.rm = TRUE)
item_means_llm   <- colMeans(res_llm$wide,   na.rm = TRUE)

broken_items <- c("NCz8AP")

item_means_human <- item_means_human[setdiff(names(item_means_human), broken_items)]
item_means_llm   <- item_means_llm[setdiff(names(item_means_llm),   broken_items)]

hist(item_means_human, main = "Distribution of % correct — humans", xlab = "% correct")
hist(item_means_llm,   main = "Distribution of % correct — LLMs",   xlab = "% correct")
```

```{r}
#| label: accuracy-join

common_items <- intersect(names(item_means_human), names(item_means_llm))

df_plot <- data.frame(
  item  = common_items,
  human = item_means_human[common_items],
  llm   = item_means_llm[common_items]
) %>%
  dplyr::left_join(df_items, by = c("item" = "main_id"))

# how many items survived into the comparison, and did any fail to match df_items?
c(
  human_items  = length(item_means_human),
  llm_items    = length(item_means_llm),
  common       = length(common_items),
  unmatched    = sum(is.na(df_plot$concept))
)
```

```{r}
#| label: accuracy-fit

fit <- lm(llm ~ human, data = df_plot)
r2  <- summary(fit)$r.squared
r2
```

### Human vs LLM accuracy

```{r}
#| label: fig-accuracy-labelled
#| fig-cap: "Item accuracy, humans vs LLMs, with item labels."

ggplot(df_plot, aes(x = human, y = llm, label = item)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "Item difficulty: human vs LLM",
    subtitle = paste0("R² = ", round(r2, 3)),
    x = "% correct (human)",
    y = "% correct (LLM)"
  ) +
  theme_minimal()

```
```{r}
# the hard for everyone's
df_plot$item[df_plot$human < 0.1 & df_plot$llm < 0.1]
# the hard for LLM's 
df_plot$item[df_plot$human > 0.6 & df_plot$llm == 0]
# the hard for Humans
df_plot$item[df_plot$human < 0.5 & df_plot$llm > 0.5]
```

### Coloured by concept

```{r}
#| label: fig-accuracy-concept
#| fig-cap: "Item accuracy coloured by concept."

ggplot(df_plot, aes(x = human, y = llm, color = concept)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "Item difficulty: human vs LLM",
    subtitle = paste0("R² = ", round(r2, 3)),
    x = "% correct (human)", y = "% correct (LLM)", color = "Concept"
  ) +
  theme_minimal()
```

### Coloured by grid size

```{r}
#| label: fig-accuracy-size
#| fig-cap: "Item accuracy coloured by grid size."

ggplot(df_plot, aes(x = human, y = llm, color = size)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "Item difficulty: human vs LLM",
    subtitle = paste0("R² = ", round(r2, 3)),
    x = "% correct (human)", y = "% correct (LLM)",
    color = "Grid size"   # was mislabelled "Concept" in the original script
  ) +
  theme_minimal()
```
### Colored by human rt

```{r}
#| label: fig-accuracy-rt
#| fig-cap: "Item accuracy coloured by human rt."


item_rt_human <- df_human %>%
  dplyr::filter(participant != 0) %>%
  dplyr::summarise(rt = mean(rt, na.rm = TRUE), .by = c(participant, item_main_id)) %>%
  dplyr::summarise(human_rt = mean(rt, na.rm = TRUE), .by = item_main_id)

df_plot <- df_plot %>%
  dplyr::select(-dplyr::any_of(c("mean_rt", "mean_rt.x", "mean_rt.y", "human_rt"))) %>%
  dplyr::left_join(item_rt_human, by = c("item" = "item_main_id"))

stopifnot("human_rt" %in% names(df_plot))
sum(is.na(df_plot$human_rt))  # now a real check


ggplot(df_plot, aes(x = human, y = llm, color = human_rt)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_viridis_c(option = "viridis"#, transform = "log10"
                        ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "Item difficulty: human vs LLM",
    subtitle = paste0("R² = ", round(r2, 3)),
    x = "% correct (human)", y = "% correct (LLM)", color = "Mean human RT"
  ) +
  theme_minimal()

```


### Accuracy by concept

```{r}
#| label: concept-long

concept_order <- df_plot %>%
  dplyr::group_by(concept) %>%
  dplyr::summarise(llm_mean = mean(llm, na.rm = TRUE)) %>%
  dplyr::arrange(dplyr::desc(llm_mean)) %>%
  dplyr::pull(concept)

df_long <- df_plot %>%
  tidyr::pivot_longer(
    cols      = c(human, llm),
    names_to  = "source",
    values_to = "pct_correct"
  ) %>%
  dplyr::mutate(concept = factor(concept, levels = concept_order))

# items per concept — uneven cells make the bars unequally precise
df_plot %>% dplyr::count(concept) %>% dplyr::arrange(dplyr::desc(n))
```

```{r}
#| label: fig-concept-bars
#| fig-cap: "Mean % correct per concept (bars) with individual items (points), ordered by LLM performance."

ggplot(df_long, aes(x = concept, y = pct_correct, fill = source)) +
  stat_summary(
    fun = mean, geom = "col",
    position = position_dodge(width = 0.8),
    width = 0.7, alpha = 0.6
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    size = 1.8, alpha = 0.7, shape = 21, color = "black"
  ) +
  scale_fill_brewer(palette = "Set2") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "% correct by concept: humans vs LLMs",
    x = "Concept", y = "% correct", fill = "Source"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Bars are means **across items**, so every item counts equally regardless of how
many respondents saw it — this is not the same quantity as pooled trial-level
accuracy. The x-axis is sorted by LLM performance, so any apparent trend in the
human bars across the axis is partly an artefact of sorting on a correlated
variable.

## Rasch (1PL) models

### Humans

```{r}
#| label: model-human

mod_1pl_h <- mirt(wide_human, model = 1, itemtype = "Rasch", SE = TRUE)


extract.mirt(mod_1pl_h, "converged")        # TRUE / FALSE
extract.mirt(mod_1pl_h, "iterations")       # stopped at the 500 ceiling?
extract.mirt(mod_1pl_h, "secondordertest")  # TRUE = genuine maximum

item_pars_h <- data.frame(coef(mod_1pl_h, IRTpars = TRUE, simplify = TRUE)$items)
hist(item_pars_h$b, main = "Item difficulty (b) — humans", xlab = "b")

itemfit(mod_1pl_h, na.rm = FALSE, fit_stats = "infit")

theta_h <- fscores(mod_1pl_h, method = "EAP")
hist(theta_h, main = "Ability (theta) — humans", xlab = "theta")
```

### LLMs

```{r}
#| label: model-llm

mod_1pl_l <- mirt(wide_llm, model = 1, itemtype = "Rasch", SE = TRUE)

extract.mirt(mod_1pl_l, "converged")        # TRUE / FALSE
extract.mirt(mod_1pl_l, "iterations")       # stopped at the 500 ceiling?
extract.mirt(mod_1pl_l, "secondordertest")  # TRUE = genuine maximum

item_pars_l <- data.frame(coef(mod_1pl_l, IRTpars = TRUE, simplify = TRUE)$items)
hist(item_pars_l$b, main = "Item difficulty (b) — LLMs", xlab = "b")

itemfit(mod_1pl_l, na.rm = FALSE, fit_stats = "infit")

theta_l <- fscores(mod_1pl_l, method = "EAP")
hist(theta_l, main = "Ability (theta) — LLMs", xlab = "theta")
```


```{r}
#| label: n-check

c(n_human = nrow(wide_human), n_llm = nrow(wide_llm))
```

## Comparing difficulty estimates

```{r}
#| label: difficulty-join

item_pars_h$item <- rownames(item_pars_h)
item_pars_l$item <- rownames(item_pars_l)

# distinct name from the accuracy section's `common_items`, which the original
# script overwrote at this point
common_items_b <- intersect(item_pars_h$item, item_pars_l$item)

df_diff <- data.frame(
  item    = common_items_b,
  b_human = item_pars_h$b[match(common_items_b, item_pars_h$item)],
  b_llm   = item_pars_l$b[match(common_items_b, item_pars_l$item)]
) %>%
  dplyr::left_join(df_items, by = c("item" = "main_id"))

nrow(df_diff)
```

```{r}
#| label: difficulty-fit

fit_b <- lm(b_llm ~ b_human, data = df_diff)
r2_b  <- summary(fit_b)$r.squared
r2_b
```

```{r}
#| label: fig-difficulty
#| fig-cap: "Rasch difficulty estimates, humans vs LLMs."

ggplot(df_diff, aes(x = b_human, y = b_llm, label = item)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
  coord_equal() +
  labs(
    title    = "Item difficulty (Rasch b): human vs LLM",
    subtitle = paste0("R² = ", round(r2_b, 3)),
    x = "Difficulty (human)",
    y = "Difficulty (LLM)"
  ) +
  theme_minimal()
```

::: callout-important
The Rasch scale is identified only up to a location shift, and the two models
were fit separately. The dashed *y = x* line therefore has no absolute meaning
here — only the *rank* correspondence between the two sets of difficulties is
interpretable without linking the scales (e.g. mean-sigma or Stocking-Lord, via
`plink` or `equateIRT`).
:::


## 2PL models

```{r}
#| label: 2PL model-human

# mod_2pl_h <- mirt(wide_human, model = 1, itemtype = "2PL")

n_items <- ncol(wide_human)

spec_prior <- mirt.model(paste0(
  "F = 1-", n_items, "\n",
  "PRIOR = (1-", n_items, ", a1, lnorm, 0.2, 0.5)"
))

mod_2pl_h <- mirt(wide_human, model = spec_prior, itemtype = "2PL", SE = TRUE,
                  technical = list(NCYCLES = 5000))

extract.mirt(mod_2pl_h, "converged")
extract.mirt(mod_2pl_h, "secondordertest")

item_pars_h <- data.frame(coef(mod_2pl_h, IRTpars = TRUE, simplify = TRUE)$items)
hist(item_pars_h$b, main = "Item difficulty (b) — humans", xlab = "b")
hist(item_pars_h$a, main = "Item discrimination (a) — humans", xlab = "a")

itemfit(mod_2pl_h, na.rm = FALSE, fit_stats = "infit")

theta_h <- fscores(mod_2pl_h, method = "EAP")
# hist(theta_h, main = "Ability (theta) — humans", xlab = "theta")

```

```{r}
item_pars_h$item <- rownames(item_pars_h)

df_ab <- item_pars_h %>%
  dplyr::left_join(df_items, by = c("item" = "main_id"))

ggplot(df_ab, aes(x = b, y = a, label = item)) +
  geom_hline(yintercept = 1, linetype = "dotted", color = "gray60") +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
  labs(
    title = "Item parameters (2PL) — humans",
    x = "Difficulty (b)",
    y = "Discrimination (a)"
  ) +
  theme_minimal()
```

```{r}
# starting values, to find which slope parameters are actually free
sv <- mirt(wide_human, model = 2, itemtype = "2PL", pars = "values")

a1_nums <- sv$parnum[sv$name == "a1" & sv$est]
a2_nums <- sv$parnum[sv$name == "a2" & sv$est]

priors <- c(
  lapply(a1_nums, function(p) c(p, "lnorm", 0.2, 0.5)),
  lapply(a2_nums, function(p) c(p, "norm",  0,   1))
)

mod_2d_h <- mirt(
  wide_human, model = 2, itemtype = "2PL",
  method   = "MHRM",
  parprior = priors,
  SE       = TRUE,
  technical = list(NCYCLES = 5000)
)

extract.mirt(mod_2d_h, "converged")
```

```{r}
theta_2d <- fscores(mod_2d_h, method = "EAP", rotate = "oblimin")
head(theta_2d)
theta_2d <- fscores(mod_2d_h, method = "EAP", rotate = "oblimin")
head(theta_2d)

df_theta <- data.frame(theta_2d)
colnames(df_theta) <- c("F1", "F2")

ggplot(df_theta, aes(x = F1, y = F2)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(alpha = 0.4, size = 1.5, color = "steelblue") +
  coord_equal() +
  labs(x = expression(theta[1]), y = expression(theta[2]),
       title = "Person ability estimates (EAP, oblimin)") +
  theme_minimal()

```

```{r}
ld <- summary(mod_2d_h, rotate = "oblimin", verbose = FALSE)$rotF
df_ld <- data.frame(item = rownames(ld), F1 = ld[, 1], F2 = ld[, 2]) %>%
  dplyr::left_join(df_items, by = c("item" = "main_id"))

ggplot(df_ld, aes(x = F1, y = F2, color = size)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(size = 2.2, alpha = 0.8) +
  labs(x = "Loading F1", y = "Loading F2", color = "Concept") +
  theme_minimal()

ggplot(df_ld, aes(x = F1, y = F2, color = concept, label = item)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(size = 2.2, alpha = 0.8) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 30,
                           show.legend = FALSE, segment.alpha = 0.4) +
  labs(x = "Loading F1", y = "Loading F2", color = "Concept") +
  theme_minimal()
```

## Session info

```{r}
#| label: session-info

sessionInfo()
```
